Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-08-14 paper.tex
stuff

Many dependent type theories support some form of inductive types. An
inductive type is given by its constructors, along with an elimination
principle which expresses that all inhabitants are constructed using
finitely many applications of the constructors.

@@

The article of Altenkirch et al. \cite{gabe} gives specification and semantics
of QIITs in a set-truncated setting. Signatures are given as lists of functors
which can be interpreted as complete categories of algebras, and completeness is
used to talk about notions of induction and recursion. However, no strict
positivity restriction is given, nor a construction of initial algebras.

@@

The logical predicate syntactic translation was introduced by Bernardy et
al. \cite{bernardy2010parametricity}. The idea that a context can be seen as a
signatures and the logical predicate translation can be used to derive the types
of induction motives and methods was described in \cite[Section
  5.3]{ttintt}. Logical relations are used to derive the computation rules in
\cite[Section 4.3]{kaposi-phd}, but only for closed QIITs. Syntactic
translations in the context of the calculus of inductive constructions are
discussed in \cite{next700}. Logical relations and parametricity can also be
used to justify the existence of inductive types in a type theory with an
impredicative universe \cite{atkey}.

@@

\noindent Now, we can reformulate induction for $\Nat$. First, we assume that
there exists a distinguished $\Nat$-algebra, named $\mathsf{InitialAlg}$. The
induction principle has the following type:
\[
 \mathsf{Induction} : (M : \mathsf{\DisplayedAlg}\,\,\mathsf{InitialAlg}) \ra \Section\, M
\]
\noindent Unfolding the definitions, it is apparent that this is the same notion
of $\Nat$-induction as we gave before. The initial algebra consists of type and value
constructors, the induction motives and methods are bundled into a displayed
algebra, and as result we get a section, containing an eliminator function together
with its $\beta$-rules. Additionally, we can define recursion using homomorphisms:
\[
\mathsf{Recursion} : (\alpha : \Alg) \ra \Morphism\,\mathsf{InitialAlg}\,\alpha
\]
This corresponds to \emph{weak initiality} in the sense of category theory: for
each algebra, there is a morphism from the weakly initial algebra to it. Strong initiality
in the setting of higher inductive types is called \emph{homotopy initiality} \cite{sojakova},
and it is defined as follows for $\Nat$:
\begin{alignat*}{5}
\mathsf{Initiality} : (\alpha : \Alg) \ra \mathsf{isContr}\,(\Morphism\,\mathsf{InitialAlg}\,\alpha)
\end{alignat*}

@@

\subsection{Interpreting Intrinsic Syntax}
Following Altenkirch and Kaposi \cite{ttintt}, one might define the syntax of a
type theory as an initial category with families (CwF) \cite{dybjer1995internal}
extended with additional type formers. The CwF part provides a calculus and
equational theory for explicit substitutions, upon which one can build
additional type structure. We present an excerpt below:

@@

So far, we were able to compute algebras, displayed algebras, morphisms and
sections, and this allows us to state recursion and induction principles.
Reiterating Section \ref{sec:general}, for a signature
$\hat{\Gamma\vdash}\,\Delta$ and assuming $\hat{\Gamma\vdash \mathsf{InitialAlg}
  : \blc{\Delta^\A}}$, we have the following types for induction and recursion:
\begin{alignat*}{5}
& \hat{\Gamma\vdash \mathsf{Induction} :
    (\gamma^D : \blc{\Delta^\D}\,\mathsf{InitialAlg})\ra \blc{\Delta^\S}\,\mathsf{InitialAlg}\,\gamma^D}\\
& \hat{\Gamma\vdash \mathsf{Recursion} :
    (\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\mathsf{InitialAlg}\,\gamma}
\end{alignat*}

>>>>

Many dependent type theories support some form of inductive types. An
inductive type is given by its constructors, along with an elimination
principle which expresses that it is enough to consider the
constructors when defining a function out of the inductive type.

@@

The article of Altenkirch et al.\ \cite{gabe} gives specification and semantics
of QIITs in a set-truncated setting. Signatures are given as lists of functors
which can be interpreted as complete categories of algebras, and completeness is
used to talk about notions of induction and recursion. However, no strict
positivity restriction is given, nor a construction of initial algebras.

@@

The logical predicate syntactic translation was introduced by Bernardy et
al.\ \cite{bernardy2010parametricity}. The idea that a context can be seen as a
signatures and the logical predicate translation can be used to derive the types
of induction motives and methods was described in \cite[Section
  5.3]{ttintt}. Logical relations are used to derive the computation rules in
\cite[Section 4.3]{kaposi-phd}, but only for closed QIITs. Syntactic
translations in the context of the calculus of inductive constructions are
discussed in \cite{next700}. Logical relations and parametricity can also be
used to justify the existence of inductive types in a type theory with an
impredicative universe \cite{atkey}.

@@

\noindent Now, we can reformulate induction for $\Nat$. First, we
assume that there exists a distinguished $\Nat$-algebra, named
$\Nat^*$. The induction principle for this algebra has the following
type:
\[
 \mathsf{Induction} : (M : \mathsf{\DisplayedAlg}\,\,\Nat^*) \ra \Section\, M
\]
\noindent Unfolding the definitions, it is apparent that this is the same notion
of $\Nat$-induction as we gave before. The initial algebra consists of type and value
constructors, the induction motives and methods are bundled into a displayed
algebra, and as result we get a section, containing an eliminator function together
with its $\beta$-rules. Additionally, we can define recursion using homomorphisms:
\[
\mathsf{Recursion} : (\alpha : \Alg) \ra \Morphism\,\Nat^* \,\alpha
\]
This corresponds to \emph{weak initiality} in the sense of category theory: for
each algebra, there is a morphism from the weakly initial algebra to it. Strong initiality
in the setting of higher inductive types is called \emph{homotopy initiality} \cite{sojakova},
and it is defined as follows for $\Nat$:
\begin{alignat*}{5}
\mathsf{Initiality} : (\alpha : \Alg) \ra \mathsf{isContr}\,(\Morphism\,\Nat^*\,\alpha)
\end{alignat*}

@@

\subsection{Interpreting Intrinsic Syntax}
Following Altenkirch and Kaposi \cite{ttintt}, one might define the
syntax of a type theory as a category with families (CwF)
\cite{dybjer1995internal} extended with additional type formers, which
supports an induction principle. The CwF part provides a calculus and
equational theory for explicit substitutions, upon which one can build
additional type structure. We present an excerpt below:

@@

So far, we were able to compute algebras, displayed algebras,
morphisms and sections, and this allows us to state recursion and
induction principles.  Reiterating Section \ref{sec:general}, for a
signature $\hat{\Gamma\vdash}\,\Delta$ and assuming $\hat{\Gamma\vdash
  \Delta^* : \blc{\Delta^\A}}$ as a candidate algebra for the HIIT, we
have the following types for induction and recursion:
\begin{alignat*}{5}
& \hat{\Gamma\vdash \mathsf{Induction} :
    (\gamma^D : \blc{\Delta^\D}\,\Delta^*)\ra \blc{\Delta^\S}\,\Delta^*\,\gamma^D}\\
& \hat{\Gamma\vdash \mathsf{Recursion} :
    (\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\Delta^*\,\gamma}
\end{alignat*}
